A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-10-13 inst/manuscript/manuscript.Rmd
section 5 edits

* Pros of Design-Based (items we are not exploring): computationally efficient, few assumptions, more naturally handles binary data,

* Pros of Model-Based (items we are not exploring): covariate inference, more efficient small-area estimation, model selection?, estimated spatial parameters to better understand spatial structure, site-by-site predictions/prediction map

>>>>

The design-based and model-based approaches to inference are fundamentally different paradigms by which to select samples and analyze data. The design-based approach incorporates randomness through sampling to estimate a population parameter. The model-based approach incorporates randomness through distributional assumptions to predict the realized value of a random variable. Though these approaches have often been compared in the literature both from theoretical and analytical perspectives, our contribution lies in studying them in a spatial context while implementing spatially balanced sampling. Aside from the theoretical differences described, a few analytical findings were particularly notable: the design decision (GRTS vs IRS) seems much more important than the analysis decision (design-based vs model-based); Independent of the analysis approach, there is no reason to prefer IRS over GRTS for spatial data -- GRTS tends to perform at least as well as IRS when there is no spatial correlation increasingly than IRS as the strength of spatial correlation increases; the gap in relative efficiency between design-GRTS and model-GRTS widens as the strength of spatial correlation increases; and when the data are skewed, interval coverage for all approaches improves both as the sample size increases and as the strength of correlation increases.

There are several benefits and drawbacks of the design-based and model-based approaches for spatial data, some of which we have not yet discussed but are worthy of consideration in future research. The design-based approach relies on few assumptions, while the model-based approach relies on rigid distributional ones. The Horvitz-Thompson estimator of means and totals is unbiased and computationally efficient, while model-based estimation of covariance parameters can be computationally burdensome, especially for likelihood-based methods such as REML that rely on inverting a covariance matrix. The design-based approach also more naturally handles binary data, free from the more complicated logistic regression formulation commonly used to handle binary data in a model-based approach. The model-based approach, however, can quantify the relationship between covariates (predictor variables) and the response variable, something the design-based approach cannot do naturally. The model-based approach also yields estimated spatial covariance parameters, which help better understand the process of study. Model selection is also possible using model-based approaches and criteria such as likelihood ratio tests or AIC [@akaike1974new]. Model-based approaches are capable of more efficient small-area estimation than design-based approaches by leveraging distributional assumptions in areas with few observed sites. Mdoel-based approaches can also compute site-by-site predictions at unobserved locations and use them to construct informative visualizations. The benefits and drawbacks of both approaches, alongside our theoretical and analytical comparisons, should be heavily considered when choosing among them. This is especially true from an analysis perspective, as we found that using a spatially balanced sampling algorithm benefits both design-based and model-based analyses.

### 2021-10-13 inst/manuscript/manuscript.tex
section 5 edits

\begin{itemize}
\item
  Pros of Design-Based (items we are not exploring): computationally
  efficient, few assumptions, more naturally handles binary data,
\item
  Pros of Model-Based (items we are not exploring): covariate inference,
  more efficient small-area estimation, model selection?, estimated
  spatial parameters to better understand spatial structure,
  site-by-site predictions/prediction map
\end{itemize}

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-barabesi2011sampling}{}%
Barabesi, L., Franceschi, S., 2011. Sampling properties of spatial total
estimators under tessellation stratified designs. Environmetrics 22,
271--278.

>>>>

The design-based and model-based approaches to inference are
fundamentally different paradigms by which to select samples and analyze
data. The design-based approach incorporates randomness through sampling
to estimate a population parameter. The model-based approach
incorporates randomness through distributional assumptions to predict
the realized value of a random variable. Though these approaches have
often been compared in the literature both from theoretical and
analytical perspectives, our contribution lies in studying them in a
spatial context while implementing spatially balanced sampling. Aside
from the theoretical differences described, a few analytical findings
were particularly notable: the design decision (GRTS vs IRS) seems much
more important than the analysis decision (design-based vs model-based);
Independent of the analysis approach, there is no reason to prefer IRS
over GRTS for spatial data -- GRTS tends to perform at least as well as
IRS when there is no spatial correlation increasingly than IRS as the
strength of spatial correlation increases; the gap in relative
efficiency between design-GRTS and model-GRTS widens as the strength of
spatial correlation increases; and when the data are skewed, interval
coverage for all approaches improves both as the sample size increases
and as the strength of correlation increases.

There are several benefits and drawbacks of the design-based and
model-based approaches for spatial data, some of which we have not yet
discussed but are worthy of consideration in future research. The
design-based approach relies on few assumptions, while the model-based
approach relies on rigid distributional ones. The Horvitz-Thompson
estimator of means and totals is unbiased and computationally efficient,
while model-based estimation of covariance parameters can be
computationally burdensome, especially for likelihood-based methods such
as REML that rely on inverting a covariance matrix. The design-based
approach also more naturally handles binary data, free from the more
complicated logistic regression formulation commonly used to handle
binary data in a model-based approach. The model-based approach,
however, can quantify the relationship between covariates (predictor
variables) and the response variable, something the design-based
approach cannot do naturally. The model-based approach also yields
estimated spatial covariance parameters, which help better understand
the process of study. Model selection is also possible using model-based
approaches and criteria such as likelihood ratio tests or AIC (Akaike,
1974). Model-based approaches are capable of more efficient small-area
estimation than design-based approaches by leveraging distributional
assumptions in areas with few observed sites. Mdoel-based approaches can
also compute site-by-site predictions at unobserved locations and use
them to construct informative visualizations. The benefits and drawbacks
of both approaches, alongside our theoretical and analytical
comparisons, should be heavily considered when choosing among them. This
is especially true from an analysis perspective, as we found that using
a spatially balanced sampling algorithm benefits both design-based and
model-based analyses.

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-akaike1974new}{}%
Akaike, H., 1974. A new look at the statistical model identification.
IEEE transactions on automatic control 19, 716--723.

\leavevmode\hypertarget{ref-barabesi2011sampling}{}%
Barabesi, L., Franceschi, S., 2011. Sampling properties of spatial total
estimators under tessellation stratified designs. Environmetrics 22,
271--278.
